How to select predictive models for decision-making or causal inference?

### 2023-12-17 how_to_select_causal_models.tex
iter

Prediction models have long been used in medicine, as with risk
score or prognostic models \cite{moons2009prognosis,steyerberg2019clinical}.
While these have historically been simple models on simple data, this is
changing with progress in machine learning and richer sources of medical data
\cite{beam2018big,rajkomar2019machine}. Health predictions can now integrate medical images
\cite{khojaste2022deep,zhang2019radiological,yala2019deep,shen2019deep,nassif2022breast},
patient records
\cite{mooney2018bigdata,desai2020comparison,simon2018predicting} or clinical
notes \cite{horng2017creating,wang2020prediction,spasic2020clinical}.
Complex data is difficult to control and model, but these models are
validated by verifying the accuracy of the prediction on left-out data
\cite{altman2009prognosis,poldrack2020establishment,varoquaux2022evaluating}.
Given a model predicting a health outcome, precision medicine would like it to
guide decisions: will an individual benefit from an intervention such as surgery
\cite{fontana2019can}? Contrasting predictions with and without the treatment
gives an answer, but statistical validity requires causal inference
\cite{snowden_implementation_2011,sperrin2019explicit,blakely2020reflection}.

@@

Indeed, concluding on the effect of a treatment is a difficult
causal-inference task, easily compromised by confounding:
spurious associations between treatment
allocation and baseline health, \eg only prescribing a drug to mild cases
\citep{hernan_causal_2020,vanderweele2019principles}.
Predictive modeling bridges to causal inference theory under the name of
\emph{outcome models} (or G-computation, 
G-formula \cite{robins_role_1986}, Q-model
\cite{snowden_implementation_2011}, conditional mean regression
\cite{wendling_comparing_2018}). 
% Controlled allocation of treatment, as in randomized controlled trials (RCTs),
% alleviate this concern. Yet most machine-learning models are trained on
% \emph{observational} data, close to real-word practice
% \cite{black1996we,hernan_methods_2021} but challenging for causal inference.
Medical statistics and epidemiology have mostly used other
causal-inference methods, modeling treatment assignment
with propensity scores \cite{rosenbaum_central_1983,austin_moving_2015,casucci2018estimating,grose_use_2020}. Outcome modeling brings
the benefit of going
beyond average effects, estimating individualized or conditional average
treatment effects (CATE), central to precision medicine.
%
For this purpose, such methods are also invaluable on randomized trials
\cite{su2018random,lamont2018identification,hoogland2021tutorial}.

>>>>

Prediction models have long been used in medicine, as with risk
score or prognostic models \cite{moons2009prognosis,steyerberg2019clinical}.
While these have historically been simple models on simple data, this is
changing with progress in machine learning and richer sources of medical data
\cite{beam2018big,rajkomar2019machine}. Health predictions can now integrate medical images
\cite{khojaste2022deep,zhang2019radiological,yala2019deep,shen2019deep,nassif2022breast},
patient records
\cite{mooney2018bigdata,desai2020comparison,simon2018predicting} or clinical
notes \cite{horng2017creating,wang2020prediction,spasic2020clinical}.
Complex data is difficult to control and model, but these models are
validated by verifying the accuracy of the prediction on left-out data
\cite{altman2009prognosis,poldrack2020establishment,varoquaux2022evaluating}.
Given a model predicting a health outcome, precision medicine would like it to
guide decisions: will an individual benefit from an intervention such as surgery
\cite{fontana2019can}? An estimate of the effect of the treatment can be
obtained by contrasting model predictions with and without the treatment,
but statistical validity requires causal inference
\cite{snowden_implementation_2011,sperrin2019explicit,blakely2020reflection}.

@@

Indeed, concluding on the effect of a treatment is a difficult
causal-inference task, as it can be easily compromised by confounding:
spurious associations between treatment
allocation and baseline health, \eg only prescribing a drug to mild cases
\citep{hernan_causal_2020,vanderweele2019principles}.
Predictive modeling bridges to causal inference theory under the name of
\emph{outcome models} (or G-computation, 
G-formula \cite{robins_role_1986}, Q-model
\cite{snowden_implementation_2011}, conditional mean regression
\cite{wendling_comparing_2018}). 
% Controlled allocation of treatment, as in randomized controlled trials (RCTs),
% alleviate this concern. Yet most machine-learning models are trained on
% \emph{observational} data, close to real-word practice
% \cite{black1996we,hernan_methods_2021} but challenging for causal inference.
Medical statistics and epidemiology have mostly used other
causal-inference methods, modeling treatment assignment
with propensity scores \cite{rosenbaum_central_1983,austin_moving_2015,casucci2018estimating,grose_use_2020}. Outcome modeling brings
the benefit of going
beyond average effects, estimating individualized or conditional average
treatment effects (CATE), central to precision medicine.
%
For this purpose, such methods are also invaluable on randomized trials
\cite{su2018random,lamont2018identification,hoogland2021tutorial}.
